---
pagetitle: 'Games Plants Play'
output: 
  html_document:
    toc: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
```

<!-- # Games Plants Play {.tabset} -->

## Evolution of Plant Height: Pairwise Invadability Plots and Coexistence of Different Strategies.

On this page, we explore the question whether two plants that differ in their height can coexist in the current model. On our way towards this goal, we get to know *Pairwise Invadability Plots (PIP)*, an important tool in modern Evolutionary Game Theory. As you will see, it seems that in the current model coexistence is not possible. On the next tab we therefore explore a modification of the plant height game. In that modified model coexistence becomes possible.

We start by repeating the functions *f(h)* and *g(h)*.

```{r}
f <- function(h) { 
  1 - h^alpha
}

g <- function(h.1, h.2) {
  P.L + (P.H - P.L)/(1 + exp(-beta*(h.1 - h.2)))
}
```

And here the parameter values we used on the previous page.

```{r}
alpha <- 3
beta <- 5
P.L <- 0.25
P.H <- 1
```

Next, the definition of the payoff function.

```{r}
w <- function(h.m, h.r) {
  f(h.m) * g(h.m, h.r)
}
```

We define the selective advantage *s(h', h)* (or invasion fitness) of a mutant with height *h'* in a population of plants with height *h* as the difference in payoff between a mutant when playing against a resident and a resident when playing against itself:

```{r}
s <- function(h.m, h.r) {
  w(h.m, h.r) - w(h.r, h.r)
}
```

A PIP is in fact simply a contour plot of the function *s(h', h)* with a contour line at height *s(h', h)=0*. By definition *s(h, h) = 0* and therefore a PIP always has a contour line along the diagonal (the 45-degree line). But as we will see, mutants *h'* different from *h* can also have zero-invasion fitness and these form a second, curved contour line in the PIP. Positive areas in the PIP correspond to combinations of mutants with trait value *h'* and residents with trait value *h* where the mutant has positive invasion fitness and can therefore establish itself in the population.

```{r}
PIP <- outer(seq(0, 1, 0.005), seq(0, 1, 0.005), s)
# the computes s(h', h) for all combinations of h' and h

par(pty="s")
filled.contour(t(PIP), levels = c(-1, 0, 1), main = "PIP", las = 2,
               xlab = expression("resident height h"),
               ylab = expression("mutant height h'"))
```

In order to determine whether a mutant *h'* will replace the resident *h* or coexist with it, we plot the mirror image of a PIP (mirrored over the 45-degree line).

```{r}
s.flipped.roles <- function(h.m, h.r) {
  w(h.r, h.m) - w(h.m, h.m)
}

PIP.mirrored <- outer(seq(0, 1, 0.005), seq(0, 1, 0.005), s.flipped.roles)

par(pty="s")
filled.contour(t(PIP.mirrored), levels = c(-1, 0, 1), main = "PIP.mirrored", las = 2,
               xlab = expression("height h"),
               ylab = expression("height h'"))
```

In order to determine if there are strategies (i.e., heights) that can coexist with each other, we superimpose the PIP with its mirror image.

```{r}
TEP <- ifelse(PIP > 0 & PIP.mirrored > 0, 1, ifelse(PIP < 0 & PIP.mirrored < 0, -1, 0))

par(pty="s")
filled.contour(t(TEP), levels = c(-1.5, -0.5, 0.5, 1.5), col = c("red", "blue", "green"),
plot.title={
  title(main = "PIP and PIP.mirrored superimposed")
  title(xlab=expression("height h"))
  title(ylab=expression("height h'"))
  abline(0, 1, lwd = 1)
})
```

In this last plot, the red area corresponds to combinations of heights *h'* and *h* where neither strategy can invade the other. In other words: whoever was first, wins. The blue areas correspond to combinations of heights where either the mutant or the resident wins. These results show that in the current model coexistence of plant species with different heights is not possible.
<!--: there always exists a unique height that will win against plants with any other height.-->

In fact there is quite a bit more one can read from this plot. These techniques involving PIPs are often referred to as *Adaptive Dynamics* or *Evolutionary Invasion Analysis*. If you want to learn more about this, Chapter 7 in the book by Hanna Kokko (2007) is a good starting point.

### TASKS

1. Evaluate the above code. Understand what the results mean. Try to understand the code as much as possible (but note that code is not relevant for the exam).

2. Repeat the above analysis for different parameter values.
    + Change $\alpha$ from 3 to 1.
    + Change $\alpha$ back to 3 but now change $\beta$ from 5 to 10.
    + Describe for each case how the value of the ESS changes. Explain why your findings makes biological sense.    


Submit your results as an html file, either by using the *compile report*-button or by producing an Rmarkdown document (if you know how to do this, you can find a very short introduction to Rmarkdown [**here**](https://clausrueffler.github.io/CodingWithR/Markdown.html)).
