---
pagetitle: 'Games Plants Play'
output: 
  html_document:
    toc: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
```

<!-- # Exercise: Games Plants Play {.tabset} -->

## A simple model for plant growth

Here, we re-analyse the model Hanna Kokko desribes in Chapter 6 of her book *Modelling for Field Biologists*. This model asks: Why do plants invest energy into growing a stem instead of investing all energy into leaf biomass?

Kokko assumes that the fitness of a plant is determined by two factors: firstly, how much leaf tissue there is in the plant (denoted *f*), and, secondly, how much each leaf is able to photosynthesize (denoted *g*). The amount of leaf tissue, *f*, can be expressed as the fraction of the plant’s above-ground biomass that is not ‘wasted’ on non-photosynthesizing tissue such as the stem. Leaf tissue is important because a plant’s photosynthesis depends completely on this tissue, and everything else such as building roots for survival, or forming seeds or fruits for reproduction, is ultimately limited by the amount of photosynthesis performed by the leaves. We make the simple assumption that the more photosynthesis, the more seeds or pollen a plant is able to produce. Therefore we can consider a plant’s fitness to be proportional to the product *fg*.

The next step is to model how plant height *h* affects *f* and *g*. Note that there seems to be a maximum for plant height around 100 m because plants ultimately depend on capillary action for getting water to the leaves, and gravity increasingly hinders this process at great vertical distances. So, *f* and *g* should be functions of *h*. First, *f* should be a monotonically decreasing function of *h* since more and more energy is invested into the growth of a stem. For the function *f(h)* Kokko suggests

$$
f(h)=1-h^\alpha,
$$

where the parameter $\alpha$ determines how fast *f* decreases with *h*. To explore this function we can plot it for different values of $\alpha$. First, let us define the function *f(h)* in R:

```{r klippy, r}
f <- function(h) {
  1 - h^alpha
}
```

Next, let us plot $f(h)$ for several different values of the parameter $\alpha$:

```{r}
alpha <- 2
curve(f(x), from = 0, to = 1, col ="blue", xlab = "plant height, h", ylab = "proportion of leaf tissue, f(h)")
alpha <- 3
curve(f(x), from = 0, to = 1, col ="orange", add = TRUE)
alpha <- 5
curve(f(x), from = 0, to = 1, col ="green", add = TRUE)
legend("bottomleft", legend = c(expression(paste(alpha, " = ", 2)),
                              expression(paste(alpha, " = ", 3)),
                              expression(paste(alpha, " = ", 5))),
       lty =1, col = c("blue", "orange", "green"))
```

Thus, the larger $\alpha$, the slower does *f* decrease with plant height *h*. Said differently, small values of $\alpha$ mean that growing taller has immediate large costs for leaf biomass, while large values allow the plant to grow quite tall before the full costs kick in. For example, a plant species whose structural tissue is mechanically very stable will need less of it, and it will have a high value for $\alpha$ as a consequence. One might therefore interpret $\alpha$ as a kind of ‘efficiency’ or ‘stability’ of the structural tissue. However, regardless of the value of $\alpha$ there are no resources left for leaves to grow when *h* reaches the maximum value *h=1*.

The shape of the function *g(h)* is more complicated because photosynthesis depends on the amount of light falling on a leaf, which depends on how high the leaf is situated relative to the leaves of neighboring plants. We here follow the same simplifying assumption made by Kokko and assume that the amount of light on the leaves of a focal plant with height $h'$ is affected by the height $h$ of its ‘neighbor’. The amount of sunlight for the focal plant is then assumed to depend on the height difference between the two plants, $h'-h$. Kokko suggests the function

$$
g(h', h)=P_L+\frac{P_H-P_L}{1+\exp\left(-\beta(h' - h)\right)}.
$$

This function looks rather complicated. To better understand it, let us plot it for different parameter values:

```{r}
g <- function(h.m, h.r) {
  P.L + (P.H - P.L)/(1 + exp(-beta*(h.m - h.r)))
}
# note that in the R code we use h.m (m for mutant) for h' and h.r (r for resident) for h
  
beta <- 10
P.L <- 0.2
P.H <- 1
curve(g(x, h.r = 0.5), from = 0, to = 1, ylim = c(0,1), col = "blue",
      xlab = expression("plant height, h'"),
      ylab = expression("photosythesis/leaf, g(h', h=0.5)"))

P.L <- 0.2
P.H <- 1
beta <- 5
curve(g(x, h.r = 0.5), from = 0, to = 1, add = TRUE, col = "orange")

P.L <- 0.4
P.H <- 0.8
beta <- 10
curve(g(x, h.r = 0.5), from = 0, to = 1, add = TRUE, col = "green")
legend("topleft", lty =1, col = c("blue", "orange", "green"),
       legend = c(expression(paste(beta, " = 10, ", P[L], " = 0.2, ", P[H], " = 1")),
                  expression(paste(beta, " = 5, ", P[L], " = 0.2, ", P[H], " = 1")),
                  expression(paste(beta, " = 10, ", P[L], " = 0.4, ", P[H], " = 0.8"))))
```

In the above code, the height $h$ of the neighboring plant, denoted $h.r$ in the code, is set equal to 0.5.

The shape of the function is S-shaped as a function of the height $h'$ of the focal plant. As $h'$ approaches 0, the value of $g(h',h)$ approaches the value of $P_L$. Thus, $P_L$ can be interpreted as the amount of photosynthesis of a plant without stem that is maximally shaded. On the other end, as $h'$ approaches 1, the value of $g(h',h)$ approaches the value of $P_H$. Thus, $P_H$ can be interpreted as the amount of photosynthesis of a plant with a stem of maximum length that experience no shading from the other plant. If both plants are equally high, $h' = 0.5 = h$, then $g(h',h)$ is exactly halfway between $P_L$ and $P_H$.

Now we have everything in place to look at the payoff of a plant with height $h'$ when competing against plants with height $h$, denoted $w(h', h)$. The payoff for a plant with height $h$ competing against plants with height $h'$ is denoted by reversing the order of the arguments in the payoff function $w$.

$$
\begin{align}
W(h', h) & = f(h')g(h', h)\\
W(h, h') & = f(h)g(h, h')
\end{align}
$$

Next we reproduce the payoff matrix shown in Table 6.1 in Kokko (2007). First, we define the payoff function.

```{r}
w <- function(h.m, h.r) { # payoff for h'
  f(h.m) * g(h.m, h.r)
}
```

We then compute the entries in the payoff matrix using exactly the same parameter values as in Kokko (2007).

```{r}
alpha <- 3
beta <- 5
h.r <- 0.5
P.L <- 0.25
P.H <- 1
```

Payoff to a plant with height $h'$ (first number in the matrix in Kokko).

```{r}
round(outer(c(0,1/3,2/3,1), c(0,1/3,2/3,1), w), digits = 3)
```

To obtain the second entry in each cell in the matrix by Kokko (which is the the payoff if the roles of who is mutant and who resident are switched) we only have to transpose the previous payoff matrix:

```{r}
t(round(outer(c(0,1/3,2/3,1), c(0,1/3,2/3,1), w), digits = 3)) # t is the command in R that produces the transpose of a matrix
```

The next step by Hanna Kokko is to ask about the *best response* of one plant to a given strategy by the other plant, that is, the height that maximizes the payoff given a certain height of the competing plant. Let us do this by allowing any height, not just a height from a set of four. Thus, we are going to reproduce figure 6.4(b) in Kokko (2007).

First, let us plot the payoff for all possible choices of $h'$ given $h = 0.5$. The strategy that corresponds to the maximum of this curve is the optimal response $h'$ given $h$. 

```{r}
curve(w(x, h.r), from = 0, to = 1, xlab = expression("height h'"), ylab = expression("payoff w(h', h = 0.5)"))
```

In the next step, we find the value of $h'$ that results in the highest possible payoff $w(h', h)$. For this we can use the R-function *optimize* (see [**here**](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/optimize) for the help page for this function).

```{r}
best.response <- function(h.r) {
  optimize(w, interval=c(0, 1), maximum = TRUE, h.r)$maximum
}

best.response(h.r = 0.5)
```

In conclusion, if plant 2 plays strategy $h = 0.5$, then plant 1 maximizes its payoff by playing $h' = 0.58$. To obtain the best response curve, we have perform the computation above for all (or at least *many*) values of $h$ between 0 and 1. After that we repeat the same procedure from the perspective plant 2.

```{r}
best.response.curve <- sapply(seq(0, 1, 0.01), best.response) # applies the function best.response to all values of h from 0 to 1 in steps of 0.01, thus, to 101 different resident strategy values

par(pty="s")
plot(seq(0, 1, 0.01), best.response.curve,
     type = "l", xlim = c(0,1), ylim = c(0,1), asp = 1,
     xlab = expression("height h of resident plant"),
     ylab = expression("height h' of mutant plant"))
lines(best.response.curve, seq(0, 1, 0.01), type = "l", lty = 2)
legend("bottomright", legend = c("best resp. to h", "best resp. to h'"), lty = c(1, 2)) 
```

What this shows, is that this game between two plants has exactly one Nash equilibrium given by the intersection of the two best response curve. Thus, if plants have approximately the height 0.62, then all plants play the best response with respect to each other. In fact, this Nash equilibrium is also an evolutionary stable strategy (ESS, the first ESS condition is fulfilled).

### TASKS

  (1)
Evaluate the above code. Understand what the results mean. Try to understand the code as much as possible (but note that code is not relevant for the exam).

  (2)
Produce the graphs *photosynthesis/leaf vs. plant height* and *payoff w(h', h) vs. height h'$ for a different value of the resident strategy h*. Select an h-value of your own choice. How do the graphs differ from those with *h=0.5*. Give an interpretation for the observed difference.

Submit your results as an html file, either by using the *compile report*-button or by producing an Rmarkdown document (if you know how to do this, you can find a very short introduction to Rmarkdown [**here**](https://clausrueffler.github.io/CodingWithR/Markdown.html)).



